Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-08-09 submission/manuscript.Rmd
draft text female specific

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022].

We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B). We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only).

- Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic cell cycle and cilium movement, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion - )
- XY only - protein localization to mitochondrion
- Shared germline DEGs
	- meiotic and 
- Shared germline genes included female gamete generation

A significant portion of germline genes were dysreguated in both male and female _Kdm5c_ mutants (Figure 1B). We then compared the log 2 fold change of shared germline DEGs to assess if sex impacts the degree of germline gene misexpression. 
	* Results of this, are females more sensitive? (Figure 1C)

We then wondered if the increased number of germline DEGs in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes. However, both shared and female-specific germline DEGs did not display a significant bias towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 1E). Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.

@@

@@

\begin{figure}
  \centering 
  \includegraphics{../../KDM5C_Germ_Dev_Sex_RA/submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Sex influences the degree of Kdm5c-KO germline gene misexpression but not the sex of germ cell-enriched genes]{\textbf{Tissue-enriched genes are misexpressed in the \textit{Kdm5c}-KO brain.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B). Left - MA plot of mRNA-sequencing. Right - Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal). Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red. \textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}. Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red. \textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}. Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
  \label{figurelabel}
\end{figure}

>>>>

It is currently unknown if the misexpression of germline genes is influenced by chromosomal sex, as previous studies on germline gene repressors have been conducted exclusively in males. We explored the impact of sex upon germline gene misexpession by comparing their dysregulation in male _Kdm5c_ hemizygous knockout (XY _Kdm5c_-KO), female homozygous knockout (XX _Kdm5c_-KO) and female heterozygous (XX _Kdm5c_-HET) EpiLCs. [@samantaActivationXistEvolutionarily2022]. We first identified differentially expressed genes (DEGs) compared to sex-matched wild-type controls (DESeq2[@loveModeratedEstimationFold2014a], log2 fold change > `r l2fcco_ESCEpiLC`, q < 0.1) and then filtered for germline-enriched genes.

Homozygous and heterozygous females expressed over double the number of germline-enriched genes than hemizygous males (Figure 4A). While the majority of germline DEGs in _Kdm5c_-KO males were also dysregulated in females (74%), there were also many male-specific and female-specific germline DEGs (Figure 4A-B). We compared the known functions of germline genes dysregulated in all samples (shared), only in females (XX only - XX _Kdm5c_-KO, XX _Kdm5c_-HET, or both), and only in males (XY only). Gene ontologies uniquely enriched in female-specific germline DEGs included meiotic (meiotic cell cycle) and flagellar (cilium movement) genes, while mitochondrial and cell signaling gene ontologies were enriched in male-specifc DEGs (protein localization to mitochondrion).

Germline genes dysregulated in both sexes were also enriched for meiotic ontologies (meiotic nuclear division), as well as egg-specific genes (female gamete generation). The majority of these shared germline DEGs had a greater log2 fold change from wild-type in females compared to males (Figure 4D-F). The increased number of germline genes and degree of dysregulation in females could be caused by improper X chromosome inactivation (XCI), as the X chromosome is enriched for many testis-specific germline genes[XXX]. However, both shared and female-specific germline DEGs were not biased towards the X chromosome, with the majority of genes lying on autosomes instead (Figure 4G). Altogether these results indicate female EpiLCs have increased sensitivity to germline gene misexpression with KDM5C loss, likely independent of potential XCI defects.

@@

- XX vs XY
- While many germline genes have roles in both the male and female germline, some display sex-biased expression or have functions unique to eggs and sperm. Many of the germline genes characterized in the male _Kdm5c_-KO brain have sperm-specific functions. It is currently unclear if sperm genes are also dysregulated in female _Kdm5c_ mutants, or if they would instead misexpress egg-biased genes.

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLC_XXvsXY.pdf}
  \caption[Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.]{\textbf{Chromosomal sex influences \textit{Kdm5c}-KO germline gene misexpression.} \textbf{A-B.} Expression of tissue-enriched genes (Li et al 2017) in the male \textit{Kdm5c}-KO amygdala (A) and hippocampus (B). Left - MA plot of mRNA-sequencing. Right - Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test \textbf{C.} Left - Average bigwigs of an example aberrantly expressed testis-enriched DEG, \textit{FK506 binding protein 6 (Fkbp6)} in the wild-type (WT) and \textit{Kdm5c}-KO (5CKO) amygdala (red) and hippocampus (teal). Right - Expression of \textit{Cyct} in wild-type tissues from NCBI Gene, with testis highlighted in blue and brain tissues highlighted in red. \textbf{D.} Left - Average bigwigs of an example ovary-enriched DEG, \textit{Zygotic arrest 1 (Zar1)}. Right - Expression of \textit{Zar1} in wild-type tissues from NCBI Gene, with ovary highlighted in teal and brain tissues highlighted in red. \textbf{E.} Left - Average bigwigs of an example liver-enriched DEG, \textit{Apolipoprotein C-I (Apoc1)}. Right - Expression of \textit{Apoc1} in wild-type tissues from NCBI Gene, with liver highlighted in orange and brain tissues highlighted in red.}
  \label{figurelabel}
\end{figure}
